Reanalysis Step 2: Multiple Imputation (Clean Version)

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
predictor matrix fixes

```{r setup-predictors}
# Use quickpred to get smart predictor selection
# This selects predictors with correlation >= 0.10 (R^2 >= 0.01)
# As recommended in Van Buuren's MICE documentation

# Variables that should always be included as predictors (key variables)
vars_always_predict <- c(
  "ever_lapse_binary",      # Exposure
  "baseline_VA_logMAR",     # Key outcome
  "outcome_VA_logMAR",      # Key outcome
  "age",                    # Demographics
  "gender_cat",
  "race_ethnic_cat",        # Demographics
  "insurance_cat",          # Demographics/SES
  "CCI", "DCSI",           # Comorbidities
  "person_dr",              # DR severity - key clinical variable
  "anti_VEGF",              # Key treatment
  "focal_laser_flag",       # Treatment
  "PRP_flag",               # Treatment
  "glaucoma_bef_hitplus_cat",  # Ocular comorbidity
  "otherretina_bef_hitplus_cat", # Ocular comorbidity
  "catsurg_before_hitplus_cat",  # Cataract surgery
  "fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr" # Fellow-eye predictors
)

@@

# Generate predictor matrix using quickpred
# Lower threshold to 0.05 to capture more predictive relationships
# This includes variables with correlations >= 0.05 (R^2 >= 0.0025)
pred_matrix <- quickpred(
  df_mice, 
  exclude = vars_never_predict,
  include = vars_always_predict,
  mincor = 0.05,  # Lowered from 0.10 to include more predictors
  method = "pearson"
)

@@

@@

>>>>

```{r setup-predictors}
# Create predictor matrix based on theoretical relationships
# Start with empty matrix (no predictors)
pred_matrix <- matrix(0, nrow = ncol(df_mice), ncol = ncol(df_mice))
colnames(pred_matrix) <- names(df_mice)
rownames(pred_matrix) <- names(df_mice)

@@

@@

# Set default predictors for imputation variables
# Demographics predict most variables
demographic_vars <- c("age", "gender_cat", "race_ethnic_cat", "insurance_cat")
for (var in c(vars_impute_continuous, vars_impute_binary, vars_impute_categorical, vars_impute_ordinal)) {
  if (var %in% rownames(pred_matrix)) {
    for (demo in demographic_vars) {
      if (demo %in% colnames(pred_matrix)) {
        pred_matrix[var, demo] <- 1
      }
    }
  }
}

# Comorbidities predict most variables
comorbidity_vars <- c("CCI", "DCSI", "glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat")
for (var in c(vars_impute_continuous, vars_impute_binary, vars_impute_categorical, vars_impute_ordinal)) {
  if (var %in% rownames(pred_matrix)) {
    for (comorb in comorbidity_vars) {
      if (comorb %in% colnames(pred_matrix)) {
        pred_matrix[var, comorb] <- 1
      }
    }
  }
}

# Treatment variables predict outcomes and DR severity
treatment_vars <- c("anti_VEGF", "focal_laser_flag", "PRP_flag", "other_inject")
for (var in c("intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")) {
  if (var %in% rownames(pred_matrix)) {
    for (tr in treatment_vars) {
      if (tr %in% colnames(pred_matrix)) {
        pred_matrix[var, tr] <- 1
      }
    }
  }
}

# Fellow-eye variables predict corresponding variables
fellow_vars <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr")
if ("baseline_VA_logMAR" %in% rownames(pred_matrix) && "fellow_baseline_VA" %in% colnames(pred_matrix)) {
  pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1
}
if ("intermediate_VA_logMAR" %in% rownames(pred_matrix) && "fellow_intermediate_VA" %in% colnames(pred_matrix)) {
  pred_matrix["intermediate_VA_logMAR", "fellow_intermediate_VA"] <- 1
}
if ("outcome_VA_logMAR" %in% rownames(pred_matrix) && "fellow_outcome_VA" %in% colnames(pred_matrix)) {
  pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1
}
if ("person_dr" %in% rownames(pred_matrix) && "fellow_dr" %in% colnames(pred_matrix)) {
  pred_matrix["person_dr", "fellow_dr"] <- 1
}

# Exposure variable predicts outcomes
if ("ever_lapse_binary" %in% colnames(pred_matrix)) {
  for (var in c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")) {
    if (var %in% rownames(pred_matrix)) {
      pred_matrix[var, "ever_lapse_binary"] <- 1
    }
  }
}

# Temporal relationships for VA variables
if ("intermediate_VA_logMAR" %in% rownames(pred_matrix) && "baseline_VA_logMAR" %in% colnames(pred_matrix)) {
  pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
}
if ("outcome_VA_logMAR" %in% rownames(pred_matrix) && "baseline_VA_logMAR" %in% colnames(pred_matrix)) {
  pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1
}
if ("outcome_VA_logMAR" %in% rownames(pred_matrix) && "intermediate_VA_logMAR" %in% colnames(pred_matrix)) {
  pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1
}

@@

# Fellow-eye VA (baseline/intermediate) also inform DR severity
if ("fellow_baseline_VA" %in% colnames(pred_matrix) && "person_dr" %in% rownames(pred_matrix)) pred_matrix["person_dr", "fellow_baseline_VA"] <- 1
if ("fellow_intermediate_VA" %in% colnames(pred_matrix) && "person_dr" %in% rownames(pred_matrix)) pred_matrix["person_dr", "fellow_intermediate_VA"] <- 1
